---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rankimp

<!-- badges: start -->
<!-- badges: end -->

Mean decrease in impurity, permutation importance, SHAP, LOCO and drop-column
importance give different answers on the same data. So does one method under a
different seed, or a different fold. Reporting a single ranking from a single
method on a single fit is standard practice and is indefensible.

`rankimp` treats every combination of method, seed, fold and model as a **judge**
expressing a ranking over the predictors, and computes the Kemeny median of the
panel — with ties, with judge weights, and with confidence sets that say which
parts of the ordering the evidence actually supports.

It builds on [`ConsRank`](https://cran.r-project.org/package=ConsRank) for the
consensus itself; the contribution is the inferential layer on top.

## Installation

Not on CRAN yet.

```r
# install.packages("pak")
pak::pak("rankimp")
```

## Usage

```{r usage}
library(rankimp)

judges <- rbind(
  permutation_seed1 = c(1, 2, 3, 4),
  permutation_seed2 = c(1, 2, 3, 4),
  shap              = c(1, 3, 2, 4),
  impurity          = c(2, 1, 3, 4)
)
colnames(judges) <- c("income", "age", "balance", "region")

consensus_rank(judges)
```

Variables the panel genuinely cannot separate come out tied, which is the thing
an average of Borda scores can never say.

## Status

Early development. The consensus itself and the conversion from importance
scores to rankings work; the ingestion, inference and heterogeneity layers are
declared, documented and not yet implemented — calling them raises an error that
says which phase they belong to.

| Phase | Content | State |
|---|---|---|
| F1 | `importance_judges()`, backends, weights | partly done |
| F2 | Consensus, ties, algorithm selection | partly done |
| F3 | Bootstrap, rank confidence sets, `prob_topk()` | declared |
| F4 | Judge clustering, plots, vignettes | declared |
| F5 | CRAN, methodological paper | — |

## Related work

* [`ConsRank`](https://cran.r-project.org/package=ConsRank) — the Kemeny median
  engine this package orchestrates.
* [`e2tree`](https://cran.r-project.org/package=e2tree) — explains a forest with
  a single tree.
* `proxima` — how the forest represents the data whose variables are ranked here.
