---
output:
  github_document:
    md_extensions: -smart
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 7,
  fig.height = 4,
  dpi = 150,
  out.width = "100%"
)
```

# rankimp

<!-- badges: start -->
[![R-CMD-check](https://github.com/agostinognasso/rankimp/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/agostinognasso/rankimp/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/agostinognasso/rankimp/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/agostinognasso/rankimp/actions/workflows/test-coverage.yaml)
[![coverage](https://img.shields.io/badge/coverage-%E2%89%A591%25-brightgreen)](https://github.com/agostinognasso/rankimp/actions/workflows/test-coverage.yaml)
[![R package version](https://img.shields.io/github/r-package/v/agostinognasso/rankimp?label=version)](https://github.com/agostinognasso/rankimp/blob/main/DESCRIPTION)
[![R >= 3.5](https://img.shields.io/badge/R-%3E%3D%203.5-blue)](https://cran.r-project.org/)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://opensource.org/licenses/MIT)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
<!-- badges: end -->

Consensus ranking of variable importance, with the uncertainty attached.

`rankimp` treats every combination of importance method, model, random seed and
resample as a **judge** casting a ranking over the predictors. It computes the
Kemeny median of that panel, keeps the ties the panel cannot break, and puts
bootstrap confidence sets around the result so that a claim about which
variables matter can be falsified rather than merely stated.

It builds on [`ConsRank`](https://cran.r-project.org/package=ConsRank) for the
consensus itself. The contribution is the inferential layer on top.

## The problem

A variable importance ranking is an estimate. It comes from an estimator with
variance, computed on one sample, under one definition of importance, and it is
almost always reported as though it were a fact.

Three things move it, and they are separable.

The **definition** moves it. Mean decrease in impurity, permutation importance,
LOCO and SHAP are not noisy measurements of one underlying quantity. They ask
different questions, and on correlated predictors they have different right
answers. Permutation importance on a fitted model asks what that model would
lose. LOCO asks what a model built without the variable would lose. Those
diverge by construction.

The **estimator** moves it. Refit the same forest under a different seed and the
ordering changes, because the ensemble is random.

The **sample** moves it. Refit on another draw from the same population and the
ordering changes again. This is the one a reader cares about when they ask
whether the finding is real.

Reporting a single ranking from a single method on a single fit collapses all
three into one number and prints it with no error bar.

## What the package does instead

Every source of importance becomes a judge, the panel is aggregated into a
Kemeny median, and the median is bootstrapped. The whole package is one line of
work, and each stage is usable on its own.

```
importance scores       importance_permutation()   importance_mdi()
      |                 importance_loco()          importance_shap()
      v
   rankings             importance_to_rank()
      |
      v
   a panel              importance_judges()   <- builds all of the above
      |
      +--> weights      judge_weights()
      v
  a consensus           consensus_rank()
      |
      +--> per judge    item_consensus()
      +--> subgroups    judge_clusters()
      v
  uncertainty           rank_confsets()
      |
      +--> probability  prob_topk()
      +--> a decision   rank_select()
```

If you already have a matrix of rankings from somewhere else, start at
`consensus_rank()` and ignore the top half.

## Installation

Not on CRAN yet. Install from GitHub:

```r
# install.packages("pak")
pak::pak("agostinognasso/rankimp")
```

## A worked example, end to end

The whole package on one problem, on the data the package ships, with the truth
known in advance so that every answer can be checked against it rather than
argued over.

### The data, and what is true in it

`applications` is 800 synthetic loan applications. Every predictor enters the
outcome with a known coefficient on the standardised scale, so a coefficient and
an effect are the same number, and the ordering a ranking ought to recover is
stored on the data frame itself.

```{r applications}
library(rankimp)

truth <- sort(attr(applications, "effects"), decreasing = TRUE)
truth
```

Seven predictors. Five drive the outcome and two enter it nowhere, so `age` and
`credit_lines` have no true ordering at all and anything that orders them is
reporting noise. Two more features were built in on purpose, because a dataset
on which every importance measure agrees would say nothing about a package for
reconciling them. `income` and `bureau_score` come from one latent
creditworthiness and correlate at 0.84, so they stand in for each other and the
credit for the signal has to be divided somehow. `prior_arrears` is the largest
effect in the data and takes only six distinct values, which is the case mean
decrease in impurity handles badly. See `?applications`.

```{r data}
library(randomForest)

set.seed(1)
fit <- randomForest(default ~ ., data = applications, ntree = 300)

set.seed(2)
folds <- rsample::vfold_cv(applications, v = 3)
```

### One ranking, and no error bar

This is the standard report: fit a model, ask it once, print what it said.

```{r single}
set.seed(1)
round(importance_permutation(fit, applications, "default", n_perm = 5), 3)

round(importance_mdi(fit), 1)
```

Read those against `truth`. Permutation gets the ordering almost right, missing
only by swapping the correlated pair. Impurity puts `prior_arrears` **fourth**
when it is the largest effect in the data, which is the low-cardinality bias
doing exactly what it is known to do. Both invent an ordering for `age` and
`credit_lines`, which have none, and nothing in either output says which part of
it is which.

### Why ranks and not scores

The two vectors above are not on a common scale, and no rescaling puts them on
one, because they answer different questions. A permutation loss of 0.130 and an
impurity decrease of 58.3 cannot be averaged.

Ranks are the coarsest thing every method can be made to agree to produce.
`importance_to_rank()` converts scores to ranks, and gives equal scores equal
ranks instead of separating them by column order.

```{r to_rank}
scores <- rbind(
  permutation = { set.seed(1); importance_permutation(fit, applications, "default", n_perm = 5) },
  mdi         = importance_mdi(fit)
)
importance_to_rank(scores)
```

The price of working in ranks is that a disagreement living entirely in the
magnitudes becomes invisible. The panel keeps the scores as well, in
`attr(J, "scores")`, and `vignette("method-disagreement")` works through a case
where two methods differ by a factor of six in the scores and produce the same
ordering.

### A panel instead of a ranking

`importance_judges()` builds the panel. It runs the requested backends over
every combination of four axes, methods, models, seeds and resamples, and
returns a `judges` object: an integer matrix with one row per judge, plus the
provenance and the raw scores.

```{r judges}
J <- importance_judges(
  fit,
  methods   = c("permutation", "mdi"),
  data      = applications,
  target    = "default",
  seeds     = 1:2,
  n_perm    = 5,
  resamples = folds
)
J
```

Two methods, two seeds and three folds is twelve judges, and they produce seven
distinct rankings between them. The split is not even: all six impurity judges
return the same ranking, and no two permutation judges return the same one. Impurity importance is a deterministic function of the fitted forest and
barely moves across folds; permutation importance is itself a random estimator,
and five shuffles on this data are not enough to pin down the middle of the
table. Nothing is misconfigured. That is the size of the noise, and a single run
of either would have hidden it.

Every axis multiplies. `seeds` measures the ensemble's own noise, `fit_list`
takes several fitted models when you want a consensus that holds across
learners, and `resamples` takes an `rset` from rsample, refitting on each
analysis set and scoring on the assessment set, so that the judges are out of
sample.

Provenance is kept, one row per judge:

```{r provenance}
attr(J, "provenance")
```

### The consensus

`consensus_rank()` returns the ranking with the smallest total Kemeny-Snell
distance to all the judges.

```{r consensus}
cr <- consensus_rank(J)
cr
```

`income` and `prior_arrears` come back tied at 2. That tie is not a rounding
artefact, it is the Kemeny median declining to order the pair the panel argues
about most: seven rankings attain the same minimum, and they place
`prior_arrears` anywhere from first to fourth.

```{r consensus_all}
cr$consensus_all
```

Reporting the first of those seven would not be neutral, because which one comes
first depends on the order of your columns. The package averages each variable's
position over the whole optimal set and re-ranks with ties, and keeps the full
set available.

Against `truth` the consensus is wrong at the top, and worth being precise
about: `prior_arrears` is the largest effect in the data and the consensus
cannot separate it from `income`, which is fourth. It gets the rest right,
including both variables that do nothing.

`tau_x` is the mean Emond-Mason agreement between the consensus and the judges.
Read it as how much agreement there was to summarise. It is not a p-value and
not a goodness of fit.

```{r plot_consensus}
library(ggplot2)
autoplot(cr)
```

The grey points are every rank the judges actually gave, with point area showing
how many judges sat there. The spread is per variable, which is what `tau_x`
cannot tell you: `debt_ratio` and the two null variables are nearly unanimous,
and `prior_arrears` is spread across four ranks.

### Who disagrees, and by how much

`item_consensus()` scores each judge against the reported consensus, worst
first.

```{r item_consensus}
item_consensus(cr)
```

`judge_weights()` turns that into votes. `by = "method"` lets you hand impurity
half a vote because you do not trust it here; `by = "reliability"` gives each
judge a weight that grows with its agreement with the rest of the panel.

```{r weights}
round(judge_weights(J, by = "reliability"), 3)
```

Look at which judges reliability weighting rewards. Every impurity judge gets
1.040 and every permutation judge gets less, because the impurity judges agree
with each other perfectly and agreement with the panel is what the weight
measures. So the weighting would amplify the six judges carrying the
cardinality bias and quieten the six that do not, and it would do it while
looking like a principled correction.

That is the warning worth taking from this section rather than the syntax. It
makes a lone dissenter quieter, and here the dissenters are the ones that are
right.

### One panel, or two panels stuck together

When agreement is low, the question is whether the disagreement is spread evenly
or whether the panel has a seam in it. `judge_clusters()` looks for the seam by
k-medians in Kemeny-Snell space, where each group's centre is the Kemeny median
of its own members and is therefore a ranking you can report.

```{r clusters}
het <- judge_clusters(J)
het
```

The seam falls exactly along the method axis, and the test rejects a single
population at `p = 0.005`. It is worth checking that it is the method and not
something else, since the panel has three axes in it:

```{r seam}
provenance <- attr(J, "provenance")
table(method = provenance$method, cluster = het$cluster)
table(fold = provenance$resample, cluster = het$cluster)
```

Six and six on the method, two and two on every fold. Reading the two group
consensuses is what makes the finding usable: permutation puts `prior_arrears`
first, where the truth puts it, and impurity puts it fourth. The panel divides,
and unlike the tie in the consensus this division is not about a detail. It is
about the largest effect in the data.

The hard part of this function is not finding groups, it is refusing to find
them. Silhouette width on its own splits a homogeneous panel of six judges 62%
of the time, because judges that rank alike sit at distance zero and score a
perfect silhouette. So the observed panel's best division in two is compared
against reference panels drawn from a single population, and `k` comes back as 1
unless that comparison rejects.

### How much of this survives another sample

`rank_confsets()` puts an interval around each variable's rank. There are two
bootstraps and they answer different questions.

`type = "judges"` resamples the panel. It measures how much the consensus
depends on which sources of importance happened to be in it, and it refits
nothing, so it is cheap.

```{r boot_judges}
set.seed(7)
rank_confsets(cr, n_boot = 500)
```

`type = "data"` resamples the rows, refits every model and rebuilds the whole
panel, once per replicate. It measures whether the ordering would survive
another dataset, which is usually the question a reader actually has. It needs a
panel built by `importance_judges()`, because it needs the recipe to rebuild.

```{r boot_data}
set.seed(7)
cb <- rank_confsets(cr, type = "data")
cb
```

The two disagree, and the disagreement is the lesson. Resampling the judges puts
`prior_arrears` in `[1, 4]` and pins `employment_yrs` to exactly 5. Resampling
the data moves `prior_arrears` to `[3, 5]`, an interval that contains neither
its consensus rank of 2 nor its true rank of 1, and widens `bureau_score` to
`[2, 6]`. An interval need not contain the rank it sits beside: the consensus is
computed once on the observed panel and the interval is computed over rebuilt
ones, and when the two disagree it is the sample talking.

This gap is not particular to the example. Measured over 300 replicates per
cell, a nominal 95% set from the data bootstrap covered between 0.966 and 0.998;
the judge bootstrap covered between 0.582 and 0.929 and reached the nominal
level in none of the six cells. Resampling a panel measures how much the methods
argue with each other, which is a real quantity and a much smaller one than
sampling variability.

```{r plot_confsets}
autoplot(cb)
```

Overlapping intervals are the honest way of saying two variables cannot be
ordered on this evidence.

### From an interval to a decision

`prob_topk()` reports how often each variable landed in the top `k` across the
replicates.

```{r topk}
prob_topk(cb, k = 4)
```

Two variables the data will not keep out of the top four, one at 0.92, and
`prior_arrears` at 0.78 despite being the largest effect there is. These
probabilities are conservative in the middle of their range and accurate at the
ends: a variable reported at 0.44 is really in the top `k` about 56% of the
time, and one reported at 0.98 is there 98% of the time. They understate rather
than overstate, which is the direction to want.

`rank_select()` keeps the variables whose entire interval clears a threshold.
This is the function to reach for when someone is going to act on the answer.

```{r select}
rank_select(cb, threshold = 3)
rank_select(cb, threshold = 5)
```

At a threshold of 3 it returns one variable. At 5 it returns four, and they are
four of the five that genuinely matter. It never returns `age` or
`credit_lines` at any threshold, which is the property that counts: the
calibration study says at most 3% of its selections are undeserved, and none at
all at the thresholds that make the strongest claim, while it selects between a
third and a half of the variables that did deserve selection.

Read a short list as "these I can defend", not as "these are the ones that
matter". If it returns nothing, that is an answer.

### What the example showed

Five things, and they are not all comfortable.

The panel refused to order the two variables that have no order, at every
threshold and in both bootstraps. It found the seam, and the seam was the
method rather than the seed or the fold. It declined to separate `prior_arrears`
from `income` rather than inventing a winner. Those three are the package
working.

The other two are the reason to keep reading past the consensus. The reported
ordering is wrong at the top, because the largest effect in the data is one that
half the panel systematically underrates and averaging did not fix. And
reliability weighting, applied without looking, would have made it worse by
giving the biased half more of the vote for being self-consistent.

None of those five statements can be made from a single ranking produced by a
single method on a single fit. The last two cannot be made from a consensus
either, without the tools that take it apart.

## The theory underneath

`vignette("theory")` is the full argument. The short version:

**The consensus is a Kemeny median.** Given `K` judges each ranking the same `p`
variables,

$$\pi^{*} = \arg\min_{\pi} \sum_{k=1}^{K} w_k \, d_{KS}(\pi, \pi_k)$$

where $d_{KS}$ is the Kemeny-Snell distance, which counts pairwise
disagreements and charges half for a pair one ranking ties and the other does
not. The reason to use this and not an average of ranks is that Kemeny and Snell
showed it is the only distance satisfying a short list of requirements a rank
aggregation ought to meet, and the median that comes from it is a Condorcet
method: if a majority of judges put A above B, so does the consensus whenever
that is consistent. An average of ranks gives no such guarantee, because an
outlier's distance enters it linearly.

**Ties are part of the answer.** The optimisation is over weak orderings, that
is rankings allowed to tie. Two variables that half the panel orders one way and
half the other have no defensible ordering, and a procedure that returns one
anyway has invented it.

**The median is often not unique**, arising in 57% to 98% of replicates in the
simulation study, so the package combines the whole optimal set rather than
taking whichever the solver returned first.

**Finding it is NP-hard.** The package solves exactly by branch and bound up to
ten variables and switches to heuristics above that. The threshold is empirical:
on tied panels of thirty judges the same exact solver took 0.010 seconds at ten
variables, 0.78 at eleven and 280 at twelve. Importance panels are the hard case
for these solvers, because the unimportant variables all tie near zero.

**Agreement is measured by Emond-Mason $\tau_x$**, not Kendall's $\tau_b$.
$\tau_b$ normalises ties in a way that breaks the correspondence with the Kemeny
distance, so maximising it and minimising $d_{KS}$ stop being the same problem.
$\tau_x$ restores it, which makes the consensus and the agreement statistic two
views of one optimisation rather than two numbers that sit near each other.

**A rank is a discrete, non-smooth functional of the data.** There is no delta
method for a quantity that jumps by a whole unit when two nearly equal scores
swap places, which is why the intervals are bootstrapped, and why which
bootstrap you ran has to be stated rather than assumed.

**The intervals are wide, and that is the finding rather than a defect.** On the
hardest simulated cell the interval spans 5.1 of 8 available ranks, and on that
same cell the point estimate recovers the exact order of the signal variables
5.3% of the time. A narrower interval would be claiming more than the data hold.

## Every exported function

| The question you are asking | The function |
|---|---|
| How important does this one method say each variable is? | `importance_permutation()`, `importance_mdi()`, `importance_loco()`, `importance_shap()` |
| I have scores; give me a ranking with honest ties | `importance_to_rank()` |
| Build me a panel from fitted models, across methods, seeds, folds, models | `importance_judges()` |
| Some judges deserve less of a vote than others | `judge_weights()` |
| What ordering do the judges agree on? | `consensus_rank()` |
| Which judges disagree with that consensus? | `item_consensus()` |
| Is this one panel, or two panels stuck together? | `judge_clusters()` |
| How much of this ordering would survive another sample? | `rank_confsets()` |
| What is the chance this variable really belongs in the top five? | `prob_topk()` |
| Which variables can I defend putting in a report? | `rank_select()` |
| Show me the panel, the intervals, the groups | `autoplot()` |

Supported engines are randomForest and ranger. Adding a third means extending
`R/engines.R` and nothing else; adding a definition of importance means one
backend in `R/judges-methods.R` and one entry in the `methods` argument.

## What has been measured

Every number quoted in this README and in the documentation comes from a script
in `inst/simulations/`, and those are meant to be re-run rather than believed.

| Claim | Measured |
|---|---|
| The data bootstrap covers at its nominal level | 0.966 to 0.998 over 300 replicates per cell |
| The judge bootstrap does not | 0.582 to 0.929, nominal reached in none of six cells |
| `prob_topk()` understates rather than overstates | 0.44 reported against 56% actual; 0.98 against 98% |
| `rank_select()` is conservative | at most 3% undeserved selections, and none at the strongest thresholds |
| `judge_clusters()` refuses fake groups | silhouette alone splits one population 62% of the time; the calibrated test does not |
| Testing two groups beats testing the best `k` | power 0.633 to 0.917 against 0.233 falling to 0.067 as judges are added |

## What this does not protect you from

Every judge in a panel of tree-ensemble methods is a tree-ensemble method.
Averaging over methods, seeds, folds and two forest implementations measures how
much the answer depends on those choices. A bias that all the judges share
passes through the consensus untouched and comes out looking like agreement.

The example above is not mild about this. `prior_arrears` is a small count and
impurity importance is biased against it, and the panel caught the problem only
because it contained a method that does not share the bias. Half a panel was
enough to see the bias and not enough to correct it, which is why the reported
consensus still cannot separate the largest effect in the data from the fourth
largest. Had every judge been an impurity judge, the consensus would have been
narrow, confident and wrong in the same direction throughout, and reliability
weighting would have called that agreement a reason for confidence.

The backends also score on the data you hand them, so they are in-sample unless
you pass a holdout set or build the panel over `resamples`. In-sample importance
rewards variables the model overfit on.

Widen the panel along the axis you are worried about. A confidence set is only
as honest as the panel it summarises.

## Documentation

Start with whichever question you have.

| Vignette | What it covers |
|---|---|
| `vignette("rankimp-intro")` | the short tour: panels, the consensus, ties, weights |
| `vignette("reference")` | every function, what it returns, and when you want it |
| `vignette("theory")` | why a Kemeny median, and what a bootstrapped rank estimates |
| `vignette("stability")` | rank confidence sets and what they were measured to cover |
| `vignette("method-disagreement")` | panels that split, and reading one that does not |
| `vignette("credit-scoring")` | all four axes end to end on one problem |
| `vignette("against-set-stability")` | how this differs from `stabm` |

## Status

Version 1.0.0. Complete end to end, and not yet on CRAN. The `stable` badge is
a statement about the interface rather than about the number of users: the
names are settled and a breaking change from here goes through a deprecation
cycle.

| Phase | Content | State |
|---|---|---|
| F1 | `importance_judges()`, backends, weights | done |
| F2 | Consensus, ties, algorithm selection | done |
| F3 | Bootstrap, rank confidence sets, `prob_topk()`, `rank_select()` | done |
| F4 | Judge clustering, plots, vignettes | done |
| F5 | CRAN, methodological paper | not started |

The inferential layer came first on purpose. `consensus_rank()` orchestrates
`ConsRank`; `rank_confsets()` orchestrates nothing, and it is the part that lets
a claim about variable importance be falsified.

## Related work

* [`ConsRank`](https://cran.r-project.org/package=ConsRank): the Kemeny median
  engine this package orchestrates.
* [`stabm`](https://cran.r-project.org/package=stabm): stability of selected
  *sets*, which is a different question from the stability of an *ordering*.
  See `vignette("against-set-stability")`.
* [`e2tree`](https://cran.r-project.org/package=e2tree): explains a forest with
  a single tree.
* `Proximum`: how the forest represents the data whose variables are ranked
  here.

## References

* Kemeny, J. G. and Snell, J. L. (1962). *Mathematical Models in the Social
  Sciences*. Ginn.
* Emond, E. J. and Mason, D. W. (2002). A new rank correlation coefficient with
  application to the consensus ranking problem. *Journal of Multi-Criteria
  Decision Analysis*, 11(1), 17-28.
* Amodio, S., D'Ambrosio, A. and Siciliano, R. (2016). Accurate algorithms for
  identifying the median ranking when dealing with weak and partial rankings
  under the Kemeny axiomatic approach. *European Journal of Operational
  Research*, 249(2).

## License

MIT, see `LICENSE`.
