---
title: "Set stability is not rank stability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Set stability is not rank stability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_stabm <- requireNamespace("stabm", quietly = TRUE)
```

The nearest neighbour of this package on CRAN is
[`stabm`](https://cran.r-project.org/package=stabm), which measures the
stability of feature *selection*: given the sets of features chosen across
resamples, how much do those sets overlap? It implements twenty-odd indices —
Nogueira, Jaccard, Kuncheva, Somol — and it does that job well.

It answers a different question from this one, and the difference is not
academic. A panel can agree perfectly on *which* variables matter and disagree
completely on *how much*. Set stability cannot see the second disagreement,
because it never looks at the order.

## Two panels

Four judges rank five variables. Both panels put the same three variables in the
top three, every time. They differ only in whether the judges agree on the order
*within* that top three.

```{r}
library(rankimp)

vars <- c("income", "age", "balance", "region", "tenure")

# The judges agree on the top-3 set, and cycle its internal order.
disordered <- rbind(
  c(1, 2, 3, 4, 5),
  c(2, 3, 1, 4, 5),
  c(3, 1, 2, 4, 5),
  c(1, 3, 2, 5, 4)
)

# The judges agree on the top-3 set, and on its internal order.
ordered <- rbind(
  c(1, 2, 3, 4, 5),
  c(1, 2, 3, 4, 5),
  c(1, 2, 3, 5, 4),
  c(1, 2, 3, 4, 5)
)

colnames(disordered) <- colnames(ordered) <- vars
```

## What set stability sees

Extract the top-3 set each judge selects, and hand those sets to `stabm`.

```{r, eval = has_stabm}
top3 <- function(J) {
  lapply(seq_len(nrow(J)), function(i) colnames(J)[J[i, ] <= 3])
}

c(
  nogueira_disordered = stabm::stabilityNogueira(top3(disordered), p = 5),
  nogueira_ordered    = stabm::stabilityNogueira(top3(ordered), p = 5),
  jaccard_disordered  = stabm::stabilityJaccard(top3(disordered)),
  jaccard_ordered     = stabm::stabilityJaccard(top3(ordered))
)
```

Every index returns 1. Perfect stability, both panels, because every judge
selected `income`, `age` and `balance`. From the point of view of set stability
these two panels are indistinguishable, and a report that stops here would say
so.

## What rank consensus sees

```{r}
cr_disordered <- consensus_rank(disordered)
cr_ordered <- consensus_rank(ordered)

c(tau_disordered = cr_disordered$tau, tau_ordered = cr_ordered$tau)
```

The panels are not indistinguishable at all. The Emond-Mason `tau_x` drops from
0.95 to 0.70, and the bootstrap confidence sets say exactly where the evidence
runs out.

```{r}
set.seed(1)
rank_confsets(cr_disordered, n_boot = 400)$confsets
```

In the disordered panel, `income`, `age` and `balance` each have a rank interval
of `[1, 3]`. They are jointly the top three and individually unordered: the data
do not support the claim that `income` beats `age`. Compare the ordered panel:

```{r}
set.seed(1)
rank_confsets(cr_ordered, n_boot = 400)$confsets
```

Here every interval is a point. The order is not an artefact of which judges
happened to be in the panel.

## Where the two agree

If the question really is "which variables should I keep", `stabm` is right and
`rankimp` agrees with it:

```{r}
set.seed(1)
cb <- rank_confsets(cr_disordered, n_boot = 400)
rank_select(cb, threshold = 3)
```

The three variables clear the bar in both panels, because `rank_select()` asks
about set membership, which is the stable thing. Use `stabm` when the selected
set is the deliverable. Use this package when the *ordering* is the deliverable
— when someone will read "income is the most important driver of default" off
the top of a plot and act on it.

## The honest summary

| Question | Tool |
|---|---|
| Do resamples select the same features? | `stabm` |
| Do they select them in the same order? | `consensus_rank()`, `item_consensus()` |
| Which orderings does the evidence support? | `rank_confsets()`, `prob_topk()` |
| Which variables am I sure belong in the top k? | `rank_select()` — and `stabm` agrees |

The overlap is real and worth stating in any paper that cites both. The gap is
that a variable importance *ranking* is the object almost every applied study
actually reports, and no set-based index can put an interval around it.
