---
title: "Robustly important variables in credit scoring"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robustly important variables in credit scoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
ready <- all(vapply(
  c("randomForest", "ranger", "rsample"),
  requireNamespace, logical(1), quietly = TRUE
))
knitr::opts_chunk$set(eval = ready)
```

A lender must be able to say which characteristics drive a decision, and must be
able to defend the claim. "Our random forest's permutation importance ranked
income first" does not survive the obvious follow-up: *would a different method,
or a different seed, or a different fold, have said the same?*

This vignette answers that question end to end, using judges drawn along all
four axes at once — method, seed, fold and model family — and it ends up
refusing to answer part of it, which is the point.

## The portfolio

Synthetic, so that the right answer is known. Default risk rises with arrears
and debt, falls with income and with time in work; age and the number of credit
lines do nothing at all.

```{r data}
set.seed(7)
n <- 800
loans <- data.frame(
  income         = round(rlnorm(n, log(38000), 0.45)),
  debt_ratio     = round(pmin(rbeta(n, 2, 5) * 1.6, 1), 3),
  age            = round(rnorm(n, 44, 12)),
  employment_yrs = round(pmax(rexp(n, 1 / 7), 0), 1),
  credit_lines   = rpois(n, 4),
  prior_arrears  = rpois(n, 1.1)
)
risk <- -3.7 +
  3.2 * loans$debt_ratio +
  0.85 * loans$prior_arrears -
  0.75 * scale(log(loans$income))[, 1] -
  0.20 * scale(loans$employment_yrs)[, 1]
loans$default <- factor(
  ifelse(rbinom(n, 1, plogis(risk)) == 1, "yes", "no"), levels = c("no", "yes")
)

mean(loans$default == "yes")
```

Each coefficient times the spread of its own variable gives the effect that a
ranking ought to recover: `prior_arrears` 0.88, `debt_ratio` 0.80, `income`
0.75, `employment_yrs` 0.20, and zero for the other two. The first three are
close enough together that no honest procedure should claim to order them.

## The panel

Two model families, two methods, three folds, two seeds: 24 judges.

```{r panel}
library(rankimp)
library(randomForest)
library(ranger)

set.seed(1)
rf_fit <- randomForest(default ~ ., data = loans, ntree = 300)
set.seed(1)
rgr_fit <- ranger(default ~ ., data = loans, num.trees = 300,
                  importance = "impurity", probability = TRUE)

J <- importance_judges(
  fit_list  = list(rf = rf_fit, ranger = rgr_fit),
  methods   = c("permutation", "mdi"),
  data      = loans,
  target    = "default",
  resamples = rsample::vfold_cv(loans, v = 3),
  seeds     = 1:2
)
J
```

Any one of those 24 rows, reported alone, would look like a result.

## What the panel agrees on

```{r consensus}
cr <- consensus_rank(J)
cr
```

The consensus recovers the true ordering exactly. It would be a mistake to stop
here: `tau_x` of 0.72 says the judges are a long way from unanimous, and the
ordering they were averaged into does not carry that.

## What survives resampling

```{r confsets}
cb <- rank_confsets(cr, n_boot = 500)
cb
```

The top three variables each have the interval `[1, 3]`. The consensus puts them
in the right order and the data cannot support it.

```{r select}
rank_select(cb, threshold = 2)
rank_select(cb, threshold = 3)
```

The first call returns nothing at all. There is no variable this portfolio can
place in the top two, and a report claiming "arrears and debt are the two
drivers" would be making up the part that matters. The second call returns all
three: *"these three characteristics drive default, ahead of everything else"*
is defensible, and any statement that orders them is not.

That is the whole argument for the package in three lines of output. A point
estimate answers a question it was never entitled to answer.

```{r plot, fig.width = 6, fig.height = 4}
autoplot(cb)
```

## Where the disagreement lives

A panel of 24 that disagrees is worth taking apart before it is averaged.

```{r clusters}
het <- judge_clusters(J)
het$k
```

Two groups. The useful question is which axis they fall along, and the panel
records the provenance of every judge:

```{r composition}
provenance <- attr(J, "provenance")
table(method = provenance$method, cluster = het$cluster)
table(model = provenance$model, cluster = het$cluster)
```

The seam is the **method**, not the model family: MDI judges sit almost entirely
in one group, permutation judges mostly in the other, while randomForest and
ranger judges are spread evenly across both. Swapping the engine changes little;
swapping how importance is defined changes the answer.

```{r centres}
het$centres
```

MDI puts `debt_ratio` first and `prior_arrears` third; permutation reverses
them. `debt_ratio` is continuous and `prior_arrears` is a small count, and mean
decrease in impurity is known to favour variables with more places to split.
The panel has rediscovered that bias, and `judge_clusters()` located it without
being told to look.

This is the cross-family robustness check worth reporting: a variable ranked
highly by both engines and both methods is substantively important; one that
ranks highly only under MDI is telling you about the estimator.

## What a consensus cannot do for you

Every judge here is a tree ensemble. Averaging over methods, seeds, folds and
two forest implementations measures how much the answer depends on those
choices — and nothing else. A bias that all 24 judges share, because they are
all built from trees, passes through the consensus untouched and comes out
looking like agreement.

The split above is the reason to believe that is not an abstract worry: the
cardinality bias was visible only because the panel contained a method that does
not share it. Widen the panel along the axis you are worried about, or the
narrow confidence set you get back is measuring the wrong thing.
