---
title: "Robustly important variables in credit scoring"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robustly important variables in credit scoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> **Status.** Placeholder. The ingestion (F1) and inferential (F3) layers are
> implemented; the case study still needs the portfolio data, and the
> gradient-boosting and penalised-regression judges need engines beyond
> randomForest and ranger (roadmap).

A lender must be able to say which characteristics drive a decision, and must be
able to defend the claim. "Our random forest's permutation importance ranked
income first" does not survive the obvious follow-up: *would a different method,
or a different seed, have said the same?*

This vignette answers that question for a credit portfolio, using judges drawn
along all four axes at once — method, seed, fold, and model family.

## The panel

```{r}
library(rankimp)

J <- importance_judges(
  fit_list  = list(rf = rf_fit, rgr = ranger_fit),
  methods   = c("permutation", "shap", "mdi"),
  data      = loans,
  target    = "default",
  resamples = rsample::vfold_cv(loans, v = 10),
  seeds     = 1:5,
  weights   = c(permutation = 1, shap = 1, mdi = 0.5)
)
```

Two model families, three methods, ten folds, five seeds: 300 judges. Any one
of them, reported alone, would be a publishable-looking result.

## What survives

```{r}
cr <- consensus_rank(J)
cb <- rank_confsets(cr, n_boot = 1000)

rank_select(cb, threshold = 10)
```

## Cross-model consensus as a robustness check

The interesting comparison is between the consensus *within* a model family and
the consensus *across* families. Variables that rank highly for every family are
substantively important. Variables that rank highly only for the gradient
boosting fit are telling you about the fit, not about credit risk.

## Connecting back to the forest

`Proximum` asks how the forest represents the borrowers; `rankimp` asks which
variables drive that representation; `e2tree` renders the answer as a single
tree. The three vignettes use the same portfolio on purpose.
