---
title: "When XAI methods disagree"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{When XAI methods disagree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> **Status.** The ingestion layer is phase F1 and the heterogeneity layer is
> phase F4. This vignette fixes the interface.

A low `tau_x` is not a failure of the consensus. It is a finding about the
methods.

When predictors are correlated, marginal importance measures and conditional
ones answer different questions, and their rankings separate. Averaging them
into a single ordering destroys exactly the information that mattered.
`judge_clusters()` looks for that separation instead of hiding it.

## Building the panel

```{r}
library(rankimp)

J <- importance_judges(
  fit_list = list(rf = rf_fit, xgb = xgb_fit),
  methods  = c("permutation", "shap", "mdi"),
  weights  = c(permutation = 1, shap = 1, mdi = 0.5)
)
```

## Is there one panel, or two?

```{r}
cr <- consensus_rank(J)
cr$tau        # low: the panel does not agree

het <- judge_clusters(J, k = NULL)
autoplot(het) # MDS of the judges in Kemeny-Snell space
```

If the clusters line up with the marginal/conditional distinction, the honest
report is two consensus rankings with an explanation, not one consensus ranking
with a caveat.

```{r}
lapply(split(seq_len(nrow(J)), het$cluster), function(rows) consensus_rank(J[rows, ]))
```

## Why this is a result and not a problem

The disagreement is diagnostic of the correlation structure of the predictors.
A panel that agrees tells you the variables are separable; a panel that splits
tells you which variables are entangled, and with what. Both are reportable.
