---
title: "How stable is a variable importance ranking?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How stable is a variable importance ranking?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> **Status.** The inferential layer is phase F3. This vignette fixes the
> interface and states what the layer has to deliver.

Two sources of instability, and they are not the same thing.

**Intrinsic.** Refit the same model on the same data with a different seed and
the ranking moves. This is variability of the *estimator*, and multi-replicate
judges measure it.

**Data.** Refit on a different fold or bootstrap sample and the ranking moves
again. This is variability of the *estimand*, and multi-fold judges measure it.

A ranking that survives the first but not the second is telling you the sample
is too small to support the claim, not that the model is unstable.

## Rank confidence sets

```{r}
library(rankimp)

J <- importance_judges(
  fit_list  = list(rf = rf_fit),
  methods   = "permutation",
  resamples = rsample::vfold_cv(df, v = 10)
)

cr <- consensus_rank(J)
cb <- rank_confsets(cr, n_boot = 500, level = 0.95)

autoplot(cb)
```

The plot shows each variable at its consensus rank with the interval of ranks it
plausibly occupies. Overlapping intervals mean the two variables cannot be
ordered on this evidence — which is the statement most importance plots refuse
to make.

## Probability of membership in the top k

```{r}
prob_topk(cb, k = 5)
```

"`income` is the third most important variable" is a claim that cannot be
falsified. "`income` is in the top five with probability 0.97" can be.

## Selection with a guarantee

```{r}
rank_select(cb, threshold = 10)
```

A variable is selected only when its whole confidence set sits above the
threshold. Selecting on the point estimate would ignore that the ranking was
estimated in the first place.
