---
title: "How stable is a variable importance ranking?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How stable is a variable importance ranking?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_rf <- requireNamespace("randomForest", quietly = TRUE)
```

Two sources of instability, and they are not the same thing.

**Intrinsic.** Refit the same model on the same data with a different seed and
the ranking moves. This is variability of the *estimator*, and multi-replicate
judges measure it.

**Data.** Refit on a different fold or bootstrap sample and the ranking moves
again. This is variability of the *estimand*, and multi-fold judges measure it.

A ranking that survives the first but not the second is telling you the sample
is too small to support the claim, not that the model is unstable.

`rank_confsets()` reaches both, and says which one it is measuring.
`type = "judges"` resamples the panel you already have: whatever variability you
built into it — seeds, folds, methods — is what the intervals reflect.
`type = "data"` goes back to the data instead, drawing the rows with replacement
and rebuilding the panel from scratch on every draw. The first is nearly free;
the second costs a refit of every model per replicate.

## A panel that does not quite agree

```{r}
library(rankimp)

judges <- rbind(
  permutation_seed1 = c(1, 2, 3, 4, 5),
  permutation_seed2 = c(1, 2, 3, 5, 4),
  permutation_seed3 = c(2, 1, 3, 4, 5),
  shap              = c(1, 3, 2, 4, 5),
  impurity          = c(1, 2, 4, 3, 5),
  loco              = c(2, 1, 3, 5, 4)
)
colnames(judges) <- c("income", "age", "balance", "region", "tenure")

cr <- consensus_rank(judges)
cr
```

`tau_x` summarises the agreement. Whether that summary is a fair one is a
separate question, and `item_consensus()` answers it by scoring every judge
against the consensus:

```{r}
item_consensus(cr)
```

A single judge far below the rest is a finding, not noise. Marginal and
conditional importance measures diverge by construction when predictors are
correlated, and the one in the minority will look like an outlier.

## Rank confidence sets

```{r}
set.seed(1)
cb <- rank_confsets(cr, n_boot = 500)
cb
```

Each variable sits at its consensus rank, with the interval of ranks it
plausibly occupies when the panel is resampled. Overlapping intervals mean the
two variables cannot be ordered on this evidence — the statement most importance
plots decline to make.

```{r, fig.width = 6, fig.height = 3.5}
autoplot(cb)
```

## Probability of membership in the top k

```{r}
prob_topk(cb, k = 2)
```

"`income` is the most important variable" is a claim that cannot be falsified.
"`income` is in the top two with the probability printed above" can be.

## Selection with a guarantee

```{r}
rank_select(cb, threshold = 3)
```

A variable is selected only when its whole confidence set clears the threshold.
Selecting on the point estimate would ignore that the ranking was estimated in
the first place. The rule is conservative on purpose: it answers "which
variables am I sure about", and `prob_topk()` quantifies the doubt about the
rest.

## Resampling the data instead of the panel

Everything above resamples the panel. To ask the other question the models have
to be refitted, and `type = "data"` does exactly that: each replicate draws the
rows with replacement, refits every model, recomputes every judge and takes the
consensus again. It needs a panel built by `importance_judges()`, which keeps
the recipe — the fits, the data, the settings — so that none of it has to be
handed over twice.

The two questions can disagree, and thin data is where they do. Eight
predictors, five of them real with deliberately close effects, three pure noise,
and eighty rows:

```{r, eval = has_rf}
set.seed(7)
n <- 80
sim <- as.data.frame(matrix(rnorm(n * 8), n, 8))
names(sim) <- paste0("x", 1:8)
sim$y <- 2 * sim$x1 + 0.60 * sim$x2 + 0.55 * sim$x3 +
  0.50 * sim$x4 + 0.45 * sim$x5 + rnorm(n, sd = 1)

set.seed(1)
forest <- randomForest::randomForest(y ~ ., data = sim, ntree = 200)

set.seed(2)
sim_panel <- importance_judges(forest, methods = c("permutation", "mdi"),
                               data = sim, target = "y", seeds = 1:3)
cr_sim <- consensus_rank(sim_panel)
cr_sim
```

Six judges — two methods on three refits — and a consensus that reads like a
clean ordering. Now put an interval around it twice:

```{r, eval = has_rf}
set.seed(3)
by_judges <- rank_confsets(cr_sim, n_boot = 500)

set.seed(3)
by_data <- rank_confsets(cr_sim, type = "data")

by_data
```

Read the table before reading on: some intervals do not contain the consensus
rank they sit beside. That is the method, not a fault. Eighty rows drawn with
replacement hold about fifty distinct ones, and a weak-but-real predictor is
harder to place on that much less information, so its bootstrap ranks drift
towards worse positions. Give a replicate all eighty distinct rows instead and
it returns the point estimate exactly — which is how bootstrap bias is told
apart from a panel rebuilt wrongly.

Then ask each the same question — which variables would you certify in the top
three? — and compare the answers:

```{r, eval = has_rf}
rank_select(by_judges, threshold = 3)
rank_select(by_data, threshold = 3)
```

Where the two lists differ, the difference is the sample talking. A variable the
panel agrees on but the data will not support is exactly the case the opening
paragraphs describe: not an unstable model, a sample too small for the claim.

Two things to know about how a data replicate is built. A `resamples` axis is
replaced by the bootstrap's own in-bag/out-of-bag split, so a panel of
`models × methods × V` judges is rebuilt with `models × methods` of them and
each replicate votes with fewer judges than the point estimate did. And refits
preserve the number of trees and `mtry` and nothing else, so a forest with a
hand-tuned `nodesize` comes back at the engine's default — true of the `seeds`
and `resamples` axes too.

## How much is a 95% set worth?

Less than 95%, and the package would rather say so than let you assume
otherwise. `inst/simulations/rank-coverage.R` measures it: eight predictors,
five of them real, 300 simulated datasets per cell, and a check of how often the
nominal 95% set contains the rank the generating coefficients imply.

For the signal variables, resampling the data covered

* 0.895 at `n = 80` and 0.949 at `n = 200` when the effects separate cleanly;
* 0.799, 0.833 and 0.892 at `n` of 80, 200 and 500 when they are close enough
  that the middle of the ranking is barely identifiable.

Resampling the judges covered less in every one of the twelve combinations
measured — 0.582 against 0.799 in the hardest cell — which is the reason
`type = "data"` exists.

The shortfall is not a defect to be tuned away. At `n = 80` with close effects
the point estimate recovered the true order of the five signal variables in 3%
of replicates: the ranking is not identifiable from that sample, and no interval
can rescue it. Read a rank confidence set as a statement about what the evidence
rules out. The rank is a discrete, non-smooth functional, and a percentile
bootstrap is not automatically valid for such things.

## A note on cost

Every bootstrap replicate solves a Kemeny problem, which is NP-hard. Exact
branch-and-bound is fast on panels that agree and pathological on panels that do
not: on tied panels of thirty judges, one exact solve took 0.010 s with ten
variables, 0.78 s with eleven, and 280 s with twelve. That last figure is a day
and a half for five hundred replicates.

`rank_confsets()` therefore resamples with the `"quick"` heuristic by default
and reports which solver it used. On twenty tied panels of ten variables,
`"quick"` returned the identical consensus and the identical `tau_x` as exact
branch-and-bound. Pass `algorithm = "exact"` if you want the guarantee inside
the bootstrap and the panel is small enough to afford it.

A data replicate adds a refit of every model to that solve, which is why
`n_boot` defaults to 50 there against 500 for the panel bootstrap, and why a run
that looks long announces its projected cost before settling in.

## Related work

`vignette("against-set-stability")` compares this machinery with
[`stabm`](https://cran.r-project.org/package=stabm), which measures whether
resamples select the same features rather than whether they order them the same
way. The two answer different questions and a careful analysis wants both.
