---
title: "How stable is a variable importance ranking?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How stable is a variable importance ranking?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Two sources of instability, and they are not the same thing.

**Intrinsic.** Refit the same model on the same data with a different seed and
the ranking moves. This is variability of the *estimator*, and multi-replicate
judges measure it.

**Data.** Refit on a different fold or bootstrap sample and the ranking moves
again. This is variability of the *estimand*, and multi-fold judges measure it.

A ranking that survives the first but not the second is telling you the sample
is too small to support the claim, not that the model is unstable.

`rank_confsets()` addresses the first directly: it resamples the judges you
already have. The second needs refitting, and so needs the ingestion layer of
phase F1.

## A panel that does not quite agree

```{r}
library(rankimp)

judges <- rbind(
  permutation_seed1 = c(1, 2, 3, 4, 5),
  permutation_seed2 = c(1, 2, 3, 5, 4),
  permutation_seed3 = c(2, 1, 3, 4, 5),
  shap              = c(1, 3, 2, 4, 5),
  impurity          = c(1, 2, 4, 3, 5),
  loco              = c(2, 1, 3, 5, 4)
)
colnames(judges) <- c("income", "age", "balance", "region", "tenure")

cr <- consensus_rank(judges)
cr
```

`tau_x` summarises the agreement. Whether that summary is a fair one is a
separate question, and `item_consensus()` answers it by scoring every judge
against the consensus:

```{r}
item_consensus(cr)
```

A single judge far below the rest is a finding, not noise. Marginal and
conditional importance measures diverge by construction when predictors are
correlated, and the one in the minority will look like an outlier.

## Rank confidence sets

```{r}
set.seed(1)
cb <- rank_confsets(cr, n_boot = 500)
cb
```

Each variable sits at its consensus rank, with the interval of ranks it
plausibly occupies when the panel is resampled. Overlapping intervals mean the
two variables cannot be ordered on this evidence — the statement most importance
plots decline to make.

```{r, fig.width = 6, fig.height = 3.5}
autoplot(cb)
```

## Probability of membership in the top k

```{r}
prob_topk(cb, k = 2)
```

"`income` is the most important variable" is a claim that cannot be falsified.
"`income` is in the top two with the probability printed above" can be.

## Selection with a guarantee

```{r}
rank_select(cb, threshold = 3)
```

A variable is selected only when its whole confidence set clears the threshold.
Selecting on the point estimate would ignore that the ranking was estimated in
the first place. The rule is conservative on purpose: it answers "which
variables am I sure about", and `prob_topk()` quantifies the doubt about the
rest.

## A note on cost

Every bootstrap replicate solves a Kemeny problem, which is NP-hard. Exact
branch-and-bound is fast on panels that agree and pathological on panels that do
not: on tied panels of thirty judges, one exact solve took 0.010 s with ten
variables, 0.78 s with eleven, and 280 s with twelve. That last figure is a day
and a half for five hundred replicates.

`rank_confsets()` therefore resamples with the `"quick"` heuristic by default
and reports which solver it used. On twenty tied panels of ten variables,
`"quick"` returned the identical consensus and the identical `tau_x` as exact
branch-and-bound. Pass `algorithm = "exact"` if you want the guarantee inside
the bootstrap and the panel is small enough to afford it.

## Related work

`vignette("against-set-stability")` compares this machinery with
[`stabm`](https://cran.r-project.org/package=stabm), which measures whether
resamples select the same features rather than whether they order them the same
way. The two answer different questions and a careful analysis wants both.
